The Evolution of Phenotypically Invariant Genetic Networks

### 2017-07-27 handnotes.tex
better apx explanation
edited paper -- added Haldane's rule to discussion, added stuff on rate of speciation, and changed the description of the linear model. Also added Kalman to the appendix.

\section*{The evolution of phenotypically invariant gene networks with implications for speciation, adaptation, and neutral variation}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
\\
}

@@

\begin{abstract}
 % \textbf{
I will outline an analytical theory to study the evolution of biological systems such as gene regulatory networks, borrowing insight and tools from control engineering, systems identification, and dynamical systems theory. I will describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe neutral evolution as a random walk over the phenotypically invariant network-space. Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift, network rewiring, and speciation.
%}
\end{abstract}

@@

  In this paper we propose a framework to study the evolution of biological systems. We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution. To begin, we focus on the evolution of an idealized population. We consider the neutral evolution of a large population, evolving in a static environment. Under these circumstances we expect the population explores the manifold of phenotypically invariant (or symmetric) gene regulatory networks.

  Presently we ask two questions, (1) holding the phenotype and environment constant for evolutionary time, how much do we expect gene network organization to drift, and (2) does this contribute to speciation, primarily via the fixation of reproductive incompatibilities?

@@

  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
   Thus we say that some function $\phi(t)$ is a phenotype where, 
  \begin{align}
    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
  \end{align}
  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity.

  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$.

  We describe the impulse response as, 
  \begin{align}
    h(t) = C e^{A t} B
  \end{align}
  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection.

  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$.

  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
  \begin{align}
    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}
  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
}

We will now lay out the model in more general terms.
Suppose that the \emph{internal state} of the system
is parameterized by the concentrations of a collection of $n$ molecular species,
$S_1, \ldots, S_n$,
and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
There are also $m$ ``input'' species, whose concentrations are determined
exogenously to the system,
and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
and $\ell$ ``output'' species, whose concentrations are denoted
$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
The output is merely a linear function of the internal state:
\begin{align*}
    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
\end{align*}
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
The rate at which the $i^\text{th}$ species is produced
is a weighted sum of the concentrations of the other species
as well as the input:
\begin{align*}
    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
\end{align*}
In matrix notation, this is written more concisely as
\begin{align} \label{eqn:lti_system}
    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
\end{align}

@@

  \section*{The Model II: Linear Evolutionary Systems}

@@

  Systems with identical external dynamics do not necessarily have identical internal dynamics. Any linear and minimal system -- minimal, informally meaning that the system's external dynamics are achieved with the fewest possible number of internal components -- has identical external dynamics up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
  Two systems, $\Sigma = \{ A, B, C \}$, and $\bar{\Sigma} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same dynamics if they are related by a change of coordinates.

  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
  In other words systems with varying genetic architectures yet identical selection pressures, environment, and phenotype.

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.

@@

@@

  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}

First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$. 
 \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}

@@

  \plr{NOT TRUE?:}
  However, the process always has the property that
  changes in $A$ accumulate at constant (mean squared) rate:
  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
  \begin{align}
      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 t + O(t^2) .
  \end{align}

  \begin{align}
    A(\tau) \xrightarrow{T} A(\tau + \epsilon)
  \end{align}
  After evolutionary time $T$, the population's gene network architecture evolves from $A(\tau)$ to $A(\tau + \epsilon)$ with a probability inversely proportional to the magnitude of $\epsilon$ and proportional to the magnitude of $T$. 
\jss{Change in $\tau$ is tracking movement of population mean, not ``macro'' evolutionary change, as interpreted previously.}
  \begin{align}
    \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
    A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  \end{align}
 % \section*{Phenotypic Invariance}

    \begin{example}[Not all minimal gene networks can drift]
      If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.

   %  \begin{align}
   %     C  = I \lor B = I \\
   %     IV^{-1} &= I \\
   %     \iff V &= I
   %   \end{align}
      \end{example}
      \begin{example}[All Non-Minimal Gene Networks can Drift]

      Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
      \begin{align}
        h(t) &= \widehat{h}(t) \iff \\ 
        CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
      \end{align}
      Any two systems with equivalent impulse responses will have equivalent phenotypes.
      \end{example}

@@

  %    \begin{figure}[H]
  %      \centering
  %      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
  %      \caption{F2s from $A(2)$ and $A(2.1)$.}
  %    \end{figure}
  %      \begin{figure}[H]
  %        \begin{tabular}{cc}
  %        \includegraphics[width=0.5\textwidth]{F2s-small} \\
  %        \includegraphics[width=0.5\textwidth]{F2s} \\
  %        \includegraphics[width=0.5\textwidth]{F2s-large}
  %        \end{tabular}
  %        \caption{F2s from $A(2)$ and $A(2.1)$.}
  %      \end{figure}
        \begin{figure}[H]
          \centering
          \begin{tabular}{cc}
            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
          \end{tabular}
          \caption{F1 (left) and F2 (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
          Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).}
        \end{figure}
    \end{example}
      \begin{example}[Not all Networks can Host Incompatibilities]
        convex sets cant have DMIs

        \begin{align*}
          h(t) = 2 e^{- \theta t}
        \end{align*}
        Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
      \end{example}

  \section*{Additionl Examples}

    \subsection*{Metabolic Network}
  \begin{align*}
    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
    \end{align*}
    \begin{align*}
      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
    \end{align*}

    \begin{align*}
      CV &= C \\
      VB &= B \\
      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
    \end{align*}
    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.

    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).

    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.

    \subsection*{Gap Gene Network}

>>>>

\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
\\
}

@@

\begin{abstract}
 % \textbf{
We introduce an analytical theory to study the evolution of biological systems such as gene regulatory networks, applying insight and tools from control engineering, systems identification, and dynamical systems theory. We describe a null model of regulatory network evolution by analytically describing the set of all linear gene networks (of any size) that produce identical phenotypes – and the evolutionary paths connecting them. Under constant selection and environmental stasis, neutral evolution proceeds as a random walk through phenotypically identical network-space. Under neutral conditions, this model can describe the rates at which Dobzhansky-Muller incompatibilities arise in reproductively isolated populations. This analysis provides insight into the mechanisms and parameters important for understanding developmental systems drift and speciation.
%}
\end{abstract}

@@

  In this paper we outline a theoretical framework to study the evolution of biological systems. We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.

  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we decribe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. In the neutral case, this set desribes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. Speciation typically occurs on timesclaes approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival. These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.
Such a system $\mathcal{S}$, is composed of two equaitons,
  \begin{align}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}
 Where the \emph{kryptotype} $\kappa(t)$, is a list of a system's molecular concentrations at time $t$ -- or more generally the systems \emph{internal dynamics}. As some system dynamics may not be visible nor relevant to selection, we distinguish the \emph{kryptotype} (as it is hidden) from the \emph{phenotype} -- which is the systems \emph{external dynamics}.
  The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment. $C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
  Thus the phenotype is simply a convolution of the system organization and the environment, 
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{A t} B u(t) dt ,
  \end{align}
  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system. As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regultory element (or promoter) for gene $i$.

%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}

@@

  \section*{Linear Evolutionary Systems}

@@

  Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.

  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
  In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.

@@

  Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

@@

%  \plr{rework this: we assume optimal phenotype is fixed and population is large but polymorphic and not strictly on the optimum}

%First we focus on a simple evolutionary scenario -- \emph{phenotypically invariant evolution}: a large population, perfectly adapted, and in a constant environment. In this circumstance we expect phenotype to be conserved throughout evolutionary time. As such we should only expect the phenotype to change as a consequence of genetic drift (small effective population size), adaptation to new selective and/or environemental pressures. These phenotypic variations should yield distinct signatures. Adaptative changes will change the optimal impulse response function $h(t) \xrightarrow{adaptation} h'(t)$. Genetic drift registers as an increase in the intrapopulation variation in $h(t)$. 
% \jss{added the phenotypic invariance stuff here (above) -- not sure if it should put elsewhere yet}

@@

  \plr{NOT TRUE?:}
  However, the process always has the property that
  changes in $A$ accumulate at constant (mean squared) rate:
  if $\calA$ is locally isomorphic to $\R^m$ around $A_0$
  (there are $m$ degrees of freedom in the solutions to \eqref{eqn:equivalence}) then
  \begin{align}
      \E\left[ \|A_t - A_0\|^2 \right] = m \sigma^2 T + O(T^2) .
  \end{align}

  %\begin{align}
  %  \Delta_{\tau} &= \left\lVert \frac{d}{d \tau} \text{vec} \left[A\left(\tau\right)\right] \right\rVert^{-1} \\
  %  A(\tau) &\xrightarrow{\mu} A(\tau \pm \mu \Delta_{\tau})
  %\end{align}
 % \section*{Phenotypic Invariance}

%    \begin{example}[Not all minimal gene networks can drift]
 %     If a gene network is minimal and all the molecular species involved in the network are under selection, such that $C$ is the $n \times n$ identity matrix  ($C = I_{n}$), the only acceptable  change of coordinate matrix is the identify matrix.

   %  \begin{align}
   %     C  = I \lor B = I \\
   %     IV^{-1} &= I \\
   %     \iff V &= I
   %   \end{align}
 %     \end{example}
  %    \begin{example}[All Non-Minimal Gene Networks can Drift]
  %    
  %    Despite the existence of a unique genetic architecture in the minimal case, there still exists an infinite number of systems with larger networks that have identical external dynamics. 
  %    \begin{align}
  %      h(t) &= \widehat{h}(t) \iff \\ 
  %      CA^{\jmath}B &= \widehat{C} \widehat{A}^{\jmath} \widehat{C} \qquad \text{for } \jmath = 0, 1, \dots
  %    \end{align}
  %    Any two systems with equivalent impulse responses will have equivalent phenotypes.
  %    \end{example}

@@

  %    \begin{figure}[H]
  %      \centering
  %      \includegraphics[width=0.5\textwidth]{A2_A2-1_F2s}
  %      \caption{F2s from $A(2)$ and $A(2.1)$.}
  %    \end{figure}
  %      \begin{figure}[H]
  %        \begin{tabular}{cc}
  %        \includegraphics[width=0.5\textwidth]{F2s-small} \\
  %        \includegraphics[width=0.5\textwidth]{F2s} \\
  %        \includegraphics[width=0.5\textwidth]{F2s-large}
  %        \end{tabular}
  %        \caption{F2s from $A(2)$ and $A(2.1)$.}
  %      \end{figure}
        \begin{figure}[H]
          \centering
          \begin{tabular}{cc}
            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
            \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
          \end{tabular}
          \caption{F1 (left) and F2 (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
          Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).}
        \end{figure}

        \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$. If intrapopulation regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
    \end{example}
 %     \begin{example}[Not all Networks can Host Incompatibilities]
 %       convex sets cant have DMIs

 %       \begin{align*}
 %         h(t) = 2 e^{- \theta t}
 %       \end{align*}
 %       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
 %     \end{example}

\section*{The rate of speciation under neutrality in allopatric populations}
   In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
   \begin{align*}
      \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
    \end{align*}
    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intrapopulation regulatory variance,
    \begin{align*}
      P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
    \end{align*}
    the population mean architecture will move at rate
    \begin{align*}
      \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
    \end{align*}
    where $T$ is time measured in generations. 
    USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
    Thus the squared difference of it's phenotype will be,
    \begin{align*}
      \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
    \end{align*}
    and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
       \begin{align*}
         \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
    \end{align*}
     It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
    \begin{align*}
      \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
      \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
    \end{align*}
    and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.

    This suggests that reproductively isolated populations can speciate rapidly -- on timescales much less than $N_{e}$ generations, depending on the specifics of the fitness function.

  \section*{Additional Examples}
